---
title: "Summaries"
date: "2019-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nycflights13)
```

dplyr summarise() function can be used to calculate counts and proportions of logical values: sum(x > 10), mean(y == 0). 

When used with numeric functions, TRUE is converted to 1 and FALSE to 0. 

> This makes sum() and mean() very useful: sum(x) gives the number of TRUEs in x, and mean(x) gives the proportion.

### Exercises

1. Using nycflights13 "flights" dataset, brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. 

```{r}
flights
select(flights, flight, dep_time,dep_delay)
select(flights, flight, arr_time, arr_delay)
select(flights, flight, dep_time, sched_dep_time)
select(flights, flight, arr_time, sched_arr_time)
```


Consider the following scenarios:

  - A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

```{r}
flights %>%
  group_by(flight) %>%
  summarize(early15 = sum(arr_delay <= -15, na.rm = TRUE)/ n(),
            late15 = sum(arr_delay >= 15, na.rm = TRUE)/ n(),
            n = n()) %>%
  filter(early15 == 0.5,
         late15 == 0.5)
```


  - A flight is always 10 minutes late.

```{r}
flights %>% 
  group_by(flight) %>% 
  summarise(late10per = n_distinct(arr_delay, na.rm = TRUE) / n(), 
            min10 = mean(arr_delay, na.rm = TRUE),
            n = n()) %>%
  filter(late10per == 1 & min10 == 10)
```

  - A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.

  - 99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r}
flights %>% 
  group_by(flight) %>% 
  summarise(early30m = sum(arr_delay <= -30, na.rm = TRUE) / n(),
            late30m = sum(arr_delay >= 30, na.rm = TRUE) / n(),
            n = n()) %>% 
  filter(early30m == 0.5 & late30m == 0.5)
```


  - Which is more important: arrival delay or departure delay?
  
It depends on situation, when you have or dont have a connection flight. In that case, both of delays are quite important because you can potentially miss flight to your final destination.  

2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).
```{r}
not_cancelled <- flights %>% 
        filter(!is.na(dep_time))

#not_cancelled %>% Sample code
  #count(dest)

not_cancelled %>% 
  group_by(dest) %>% 
  summarise(n = n())

#not_cancelled %>% count(tailnum, wt = distance) Sample code

not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(n = sum(distance))
```


3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?

Because if a flight didn't leave then it was cancelled. If the condition `is.na(dep_delay)` is met, then the flight was cancelled.

4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?
```{r}
flights %>%
  group_by(day) %>%
  summarise(cancelled = mean(is.na(dep_delay)),
            mean_dep = mean(dep_delay, na.rm = T),
            mean_arr = mean(arr_delay, na.rm = T)) %>%
  ggplot(aes(y = cancelled)) +
  geom_point(aes(x = mean_dep), colour = "red") +
  geom_point(aes(x = mean_arr), colour = "blue") +
  labs(x = "Average delay per day", y = "Cancelled flights per day")
```

5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))
```{r}
flights %>%
  group_by(carrier) %>%
  summarise(dep_max = max(dep_delay, na.rm = T),
            arr_max = max(arr_delay, na.rm = T)) %>%
  arrange(desc(dep_max, arr_max))
```

6. What does the sort argument to count() do. When might you use it?
When you want to sort the cases based on the count.
